---
title: "ISE Colombia"
subtitle: "Growth rate of the leading economic indicator ISE"
author: "Miguel Martinez"
date: today
format: 
  html:
    toc: true
    theme: cosmo
editor: visual
execute:
  warning: false
---

This document consists of the process of constructing the graph of the evolution of the Indice del Seguimiento a la Economía (ISE). This is an index that yields the evolution of the economic activity in the Colombian economy in the short run. You can find the data set [here](https://www.dane.gov.co/index.php/estadisticas-por-tema/cuentas-nacionales/indicador-de-seguimiento-a-la-economia-ise).

The first step is to load the following libraries

```{r}

library(tidyverse)
library(readxl)

```

Then, we load the data:

## Load data

```{r}

# Original data

df1 = readxl::read_excel("anex-ISE-9actividades-mar2025.xlsx",
                         sheet = "Cuadro 1",
                         range = "A42:IJ57",
                         col_names = FALSE)

# Data adjusted for seasonal and calendar effects

df2 = readxl::read_excel("anex-ISE-9actividades-mar2025.xlsx",
                         sheet = "Cuadro 2",
                         range = "A11:IJ26",
                         col_names = FALSE)

```

Keep in mind that original data is already in annual percentage change. But data adjusted for seasonal and calendar effects are in levels, so we need to compute the pct_change.

## Indexing

```{r}

# Original data

df1[1, 1] = "Año"

df1[2, 1] = "Mes"

# Data adjusted for seasonal and calendar effects

df2[1, 1] = "Año"
df2[2, 1] = "Mes"

```

## **Lengthening Data**

```{r}

# Original data

df1 = df1 |>
  tidyr::pivot_longer(cols = 14:244,
                      values_to = "value") |> 
  tidyr::fill(value) |> 
  tidyr::pivot_wider(names_from = 1,
                     values_from = value) |>
  dplyr::select(Año:`Indicador de Seguimiento a la Economía`)

```

```{r}

# Data adjusted for seasonal and calendar effects

df2 = df2 |>
  tidyr::pivot_longer(cols = 2:244,
                      values_to = "value") |> 
  tidyr::fill(value) |> 
  tidyr::pivot_wider(names_from = 1,
                     values_from = value) |>
  dplyr::select(Año:`Indicador de Seguimiento a la Economía`)

```

## Merging Dataframes

```{r}

# Now we merge both dataframes: original and data adjusted for seasonal and calendar effects by (año and mes)

data = df1 |> 
  dplyr::inner_join(df2,
                    by = c("Año", "Mes"),
                    suffix = c(".Original", ".Ajustado"))

data
```

## Dtype transformation

```{r}

# First notice that some months have strings, so we need to replace them

año = data |>
  dplyr::distinct(Año)

año$Año

```

```{r}

# Then, we replace the strings next to the years 2023, 2024 and 2025

data = data |>
  dplyr::mutate(
    Año = stringr::str_remove_all(Año, "[pr]"))

```

```{r}

# Now we transform the "string" format of months to "numeric"

data = data |>
  dplyr::mutate(
    Mes = recode(Mes,
                 "Enero" = 01, "Febrero" = 02,
                 "Marzo" = 03, "Abril" = 04,
                 "Mayo" = 05, "Junio" = 06,
                 "Julio" = 07, "Agosto" = 08,
                 "Septiembre" = 09, "Octubre" = 10,
                 "Noviembre" = 11, "Diciembre" = 12))

```

```{r}

# Now we transform the remaining variables to numeric

data = data |> 
  dplyr::mutate_at(vars(1, 4:30), as.numeric)

glimpse(data)
```

## Percentage change

```{r}

# Recall that adjusted data for seasonal and calendar effects are in levels
# So we need to compute the percentage_change

data = data |>
  dplyr::mutate(
    `Indicador de Seguimiento a la Economía.Ajustado` = (
      `Indicador de Seguimiento a la Economía.Ajustado` / lag(`Indicador de Seguimiento a la Economía.Ajustado`, n = 12) - 1) * 100)

```

## Select and filtering data

```{r}

# Here we are selecting the required series and filtering from 2022

data = data |> 
  dplyr::select(Año, Mes,
                `Indicador de Seguimiento a la Economía.Original`,
                `Indicador de Seguimiento a la Economía.Ajustado`) |> 
  dplyr::filter(Año >= 2022)

glimpse(data)
```

## Reshape data

```{r}

# Now we reshape the data to the long or stacked format:

data = data |>
  tidyr::pivot_longer(cols = 3:4,
                      names_to ="Serie ISE",
                      values_to = "value")

data
```

## Graph

```{r}

zero = 0  # Baseline

ggplot(data) +
  geom_point(aes(x = Mes, y = value, colour = `Serie ISE`)) +
  geom_line(aes(x = Mes, y = value, colour = `Serie ISE`)) +
  geom_hline(yintercept = zero, color = "black", linetype = 2) +
  facet_grid(. ~ Año) +
  scale_x_continuous(breaks = seq(1,12, by = 1)) +
  scale_y_continuous(breaks = seq(-4,18, by = 2)) +
  theme_classic() +
  labs(title = "Tasa de crecimiento anual del Indicador de Seguimiento a la Economía (ISE)",
       subtitle = "Serie original y serie ajustada por efecto estacional y calendario
       2022 - 2025pr (marzo)",
       x = "Meses",
       y = "Variación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, ISE") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom")

```

## Conclusion

As you can see, both indexes (original and adjusted by stationary and calendar effects) are capturing the development of the real economic activity in the short run in Colombia. The year 2022 was exceptional, but if we see the long picture, we can acknowledge that the potential growth of the national economy is around 3%.
